---
title: "Final"
author: "Sara Gonzalez"
date: June 16, 2023
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
execute:
  message: false
  warning: false
---

# Problem 1

## Introduction

Disturbance is a major factor determining the plant community structure of ecological systems. Kangaroo rat (*Dipodomys spectabilis*) mounds are small-scale disturbances that support unique plant communities compared to surrounding habitats. These disturbances can affect the seed bank, an overlooked part of plant communities and an important trait of some plant species that allows for their persistence in unpredictable environments (Koontz & Simpson, 2013). Many might care about these seed bank disturbances because *D. spectabilis* manages the structure of microbial communities (Herrera et al., 2001). These microbes have the ability to recycle the primary elements that make up all living systems, especially carbon, nitrogen, and oxygen, three elements that are vital to not only plant communities, but also to animals and humans (Gupta et al., 2016). It could be useful to find how total seed number differs between kangaroo rat mound locations because it could demonstrate a relationship that tells us more about how a plant community is doing in a certain rat mound location as well as whether it is doing better or worse than other plant communities in other rat mound locations. The question I am trying to answer is: How does total seed number differ between kangaroo rat mound locations? The null hypothesis is that total seed number doesn't differ between kangaroo rat mound locations. The alternative hypothesis states the opposite, that total seed number does differ between kangaroo rat mound locations.

## Methods

### Set-up: Loading Needed Packages

```{r load-packages}
library(MASS) # have to read this in before tidyverse 
library(lme4)
library(glmmTMB) # ok if you don't have this - just comment it out

# diagnostics and model info
library(DHARMa)
library(MuMIn)
library(ggeffects)
library(lmtest) 
library(broom)
# other packages
library(tidyverse)
library(here)
library(janitor)
library(ggeffects)
library(performance)
library(naniar) 
library(flextable) 
library(car)
library(broom)
library(dplyr)
library(corrplot)
library(AICcmodavg)
library(GGally)
```

### Data Collection + Organization

```{r read-in-data}
 # data was originally a txt file. wouldn't convert to csv properly
inUrl1  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-sev/208/102459/a063411c03e4baef6707862466685442" 
infile1 <- tempfile()
try(download.file(inUrl1,infile1,method="curl"))
if (is.na(file.size(infile1))) download.file(inUrl1,infile1,method="auto")
 seeds <-read.csv(infile1,header=F 
          ,skip=1
            ,sep=","  
        , col.names=c(
                    "mnd",     
                    "dir",     
                    "loc",     
                    "species",     
                    "seeds"    ), check.names=TRUE) %>% 
   select(loc, seeds) # select columns of interest
```

The sampling methods involved the random selection of 10 rat mound locations out of 25 total. Once mounds were selected they were marked with their given number and soil samples were taken. Processing of samples Soil samples were dried in an oven for 48 h at 50 degrees Celsius and were sifted to capture small seeds and exclude large particles. After being dried, sub-samples were taken and all seeds were tested for viability and counted (Herrera et al., 2001). The data was downloaded directly from the internet and was organized by selecting columns of interest: loc (rat mound location) and seeds (seed count).

### Missing Data

```{r missing-data-visualization}
gg_miss_var(seeds) +
  theme_classic() +
  labs(title = "Missing Data for Seeds Data Set", caption = "There are no missing data observations for rat mound locations and seed count.") # added for more informative missing data plot
```

For the data set with columns of interest, there was no missing data observations. No steps to remove missing values are necessary.

### Tests

```{r anova}
ggplot(data = seeds, aes(sample = seeds)) +
stat_qq() +
stat_qq_line(color = "red") + # compares data to normal distribution
facet_wrap(~loc)

bartlett.test(seeds ~ loc, data = seeds) # test for equal variances

oneway.test(seeds ~ loc, data = seeds, var.equal = FALSE) #Welch's anova, does not need equal variance. 

seeds_anova <- seeds %>% # standard anova for post-hoc test to see which groups are different
  aov(formula = seeds ~ loc)

TukeyHSD(seeds_anova) # post-hoc test
```

As stated earlier, the null hypothesis is that total seed number doesn't differ between kangaroo rat mound locations while the alternative hypothesis is that total seed number does differ between kangaroo rat mound locations. To test the null hypothesis, we used analysis of variance (ANOVA). While it may not be the best to use for discrete data, it is good to use to determine if there is a significant difference between groups which is what we tried to do in this problem. We checked the assumptions using a qqplot to check normality as well as a Bartlett test to check homogeneity of variances, and found that the data follows neither of the assumptions. However, we were able to use a Welch's one-way ANOVA because such test does not need equal variance (Dag et al., 2018). Although the normality assumption was not met, ANOVA is robust to normal distribution, so we decided to move forward. A classic ANOVA was used anyway in order to use Tukey Honest Significant Difference (HSD) to determine which groups are different from one another.

## Results 

```{r visual}
ggplot(data = seeds, aes(x = loc, y = seeds, color =
loc, shape = loc), na.rm = FALSE) +
# jitter showing underlying data for summary statistics
geom_jitter(data = seeds, aes(x = loc, y = seeds),
alpha = 0.3, na.rm = FALSE) +
scale_color_manual(values = c("B" = "darkgreen", "D" =
"blue", "E" = "orange", "I" = "brown")) +
  labs(x = "Location", y = "Number of Seeds", title = "Total Seed Numbers in Different Kangaroo Rat mound locations", caption = "Location B has a significantly higher seed count than the others. 
       D is somewhat higher, and E & I have similar seed counts.") +
  theme_classic()
```

Using Welch's one-way ANOVA, we obtained a p-value of less than 0.001, meaning that we were able to reject the null hypothesis, and state that total seed number does differ between kangaroo rat mound locations (F(3, 716) = 14.99, p \< 0.001, and alpha = 0.05). Using the Tukey HSD test, we found that location B was significantly different from locations D, E, and I. This is consistent with what is seen in the plot. While the plot showcases some difference in location D, we found that it is not a significant difference, and that locations D, E, and I did not significantly differ from one another using the Tukey HSD test. Overall, kangaroo rat mound location B significantly differs in total seed count, while D, E, and I do not.

## References

Dag, O., Dolgun, A., Konar, N.M. (2018). onewaytests: An R Package for One-Way
 Tests in Independent Groups Designs. The R Journal, 10:1, 175-199.

Gupta, A., Gupta, R., & Singh, R. L. (2016). Microbes and environment. *Principles and Applications of Environmental Biotechnology for a Sustainable Future*, 43--84. <https://doi.org/10.1007/978-981-10-1866-4_3>

Herrera, J., Ensz, K. L., & Wilke, A. L. (2001). Stacking of seeds changes spoilage dynamics of food caches of the banner-tailed kangaroo rat (Dipodomys spectabilis). *Journal of Mammalogy*, *82*(2), 558--566. <https://doi.org/10.1093/jmammal/82.2.558>

Koontz, T. L., & Simpson, H. L. (2010). The composition of seed banks on kangaroo rat (Dipodomys spectabilis) mounds in a chihuahuan desert grassland. *Journal of Arid Environments*, *74*(10), 1156--1161. <https://doi.org/10.1016/j.jaridenv.2010.03.008>

\

# Problem 2

## Introduction

Climate-change induced alterations in environmental conditions in the alpine tundra has led to "shrubification," the expansion of woody shrubs. Shrubification is thought to change microclimatic conditions that could lead to changes in plant community composition, including seed count and weight (Seaver, 2022). Both of these factors, especially seed weight, are significant because seed weight is generally assumed to be an ecologically important life history trait in plants because it influences both dispersal ability and seedling establishment (Gross & Kromer, 1986). Additionally, seed count can be a way to determine fitness and reproductive success (Zimmerman, 1980). It could be useful to find how certain characteristics vary seed count because they could impact reproductive success in plant communities. If certain factors make reproductive success decline, it could bring harm to impacted plant communities especially those containing endangered species. The questions I am trying to answer are how seed count varies with plot type (shrub or open), plant species, and total number of inflorescences as well as what is the simplest model that explains seed count. The null hypothesis being tested here is that none of these factors impact seed count, while the alternative hypothesis is that at least one of the variables has an impact on seed count.

## Methods

### Set-up: Loading Needed Packages 

This was completed in problem 1.

### Data Collection + Organization

```{r read-data}
seedcount <- read_csv(here("data", "knb-lter-nwt", "shrubstudy_seed_ctwt.ms.data.csv")) %>% 
  clean_names() %>% # to make column names cleaner
  select(treatment, species, total_nr_infl, nr_seeds) # selecting columns of interest
```

The sampling methods involved collecting seeds from 5 different species. Seed pods were only collected when they were deemed \"mature\" - meaning they were brown and the seeds were ready to pop out. Seeds were put in a coin envelope that was taped at the corners to ensure that no seeds were lost. Envelopes were marked with the species, toothpick color, shrub number, plot number, number of inflorescences collected, and date of collection, then kept in the refrigerator until the seeds were ready to be counted. The data was organized by cleaning the names and selecting 4 columns of interest: treatment (plot type - shrub or open), species, total_nr_infl (total number of inflorescences), and nr_seeds (seed count).

### Missing Data

```{r missing-data-visual}
gg_miss_var(seedcount) +
  labs(caption = "There are no missing values for treatment, species, and total inflorescences, but a significant number for seed count.") +
  theme_classic()
```

```{r drop-NA-subset}
seedcount_subset <- seedcount %>% 
  drop_na(nr_seeds) # get rid of NA values
```

There is no missing data for the variables species, total number of inflorescences, and treatment (plot type - shrub or open). There are a considerable amount of missing observations in the variable regarding the total number of seeds. Using the drop_na function in the tidyr package, we got rid of those missing values (Wickham et al., 2023).

### Variable Relationships

```{r correlation-plot}
seedcount_cor <- seedcount_subset %>%
  select(total_nr_infl, nr_seeds) %>% 
  cor(method = "pearson")
corrplot(seedcount_cor,
         # change shape of what's in the cells
         method = "ellipse",
         addCoef.col = "black",
         mar=c(1,1,3,1))
title(main = "Correlation Plot",
         sub = "The numerical variables have a strong correlation with one another.")
```

```{r pairs-plot}
seedcount_pairs <- seedcount_subset %>% 
  select(treatment:nr_seeds)
ggpairs(seedcount_pairs) + # creates pairs plot
  labs(title = "Relationships Between Variables", caption = "There is some sort of relationship that all the variables have with each other.")
```

The Pearson correlation was used to measure the strength of the linear relationship between the variables. It has a value between -1 to 1, with a value of -1 meaning a total negative linear correlation, 0 being no correlation, and 1 meaning a total positive correlation. The variables don't have a total linear correlation, but do have a very strong positive correlation, almost totally linear at a value of 0.92. Based on the pairs plot, the variables do share some sort of relationship for the most part. One noticeable relationship is that the total number of inflorescences and total seed count has a strong positive relationship.

### Models + Choosing a Model

```{r models}
# linear model, we know this is wrong, not continuous variable (discrete)
scmod1 <- lm(nr_seeds ~ treatment + species + total_nr_infl, data = seedcount_subset)

# generalized linear model with Poisson distribution; assumption : variance = mean
scmod2 <- glm(nr_seeds ~ treatment + species + total_nr_infl, data = seedcount_subset, family = "poisson")

# generalized linear model with negative binomial distribution; glm.nb assumes nb dist.; appropriate for over dispersion, variance > mean
scmod3 <- glm.nb(nr_seeds ~ treatment + species + total_nr_infl, data = seedcount_subset)
```

```{r model-residuals}
# check diagnostics, use darma b/c GLM don't follow same assumptions, simulation of residuals
simulateResiduals(scmod1, plot = TRUE)
simulateResiduals(scmod2, plot = TRUE) # do NOT meet the assumptions at all. significant deviation, residuals are over-dispersed
simulateResiduals(scmod3, plot = TRUE) # deviation but definitely looks better than other models
```

```{r choosing-model}
MuMIn::model.sel(scmod1, scmod2, scmod3) # 100% chance of best model = 1 (weight)
```

To determine if/how seed count varies with plot type (shrub or open), plant species, and total number of inflorescences, we fit multiple models. To start, we created a linear model. We were aware this is wrong because the data we worked with were not continuous variables (discrete), but used it for comparison to better models. The second model we used was a generalized linear model (GLM) with Poisson distribution. The assumption here was that the variance and mean are equal. The third model we used was a GLM with negative binomial distribution. The assumption here was that the variance is greater than the mean. We decided to use these models because generalized linear models tend to better suit discrete data compared to linear models.

We checked the assumptions of all three of these models via simulation of residuals function in the DHARMa package (Hartig, 2022). The linear model met neither of the assumptions and had significant outliers and deviation. The Poisson distribution model met neither of the assumptions either, and was in fact worse than the linear model in that it also had over-dispersion. The negative binomial distribution model met most of the assumptions, but quantile deviations were detected. However, it definitely looked better than the other models, and other distributions did not fit with the of numerical data we had in the sense that the data was not within a certain range. Given this information, we decided to move forward by comparing the models.

We selected a model using the model selection function from the MuMin package (Barton, 2023). This takes the best model out of the three models using Akaike\'s Information Criterion (AIC), which looks for the simplest model that explains the most variance, and gives it a weight of 1, meaning that there's 100% chance of it being the best model out of the models being compared.

## Results

```{r summaries}
# model object
scmod3

# summary (z value = test stat based on z dist.)
summary(scmod3)
# species GEUROS, KOBMYO, MINOBT not significant when keeping everything else constant

confint(scmod3) # confidence intervals

# adjusted R2
r.squaredGLMM(scmod3) # only look at delta, amount of variance = 68%
```

```{r visualization}

plot(ggpredict(scmod3, terms = c("treatment", "total_nr_infl", "species")), add.data = TRUE)+
  theme_classic()+
  labs(y = "Seed Count", x = "Treatment", title = "Seed Count Based on Total Inflorescences, Species, and Treatment", caption = "The predicted seed count seems to be similar regardless of cover, a significant 
       effect is seen with total inflorescences in the TRIDAS species.") # more finalized plot
```

Based on the model comparison, we found that the third model including all variables using negative binomial distribution was the best model for looking at how seed count varies with plot type (shrub or open), plant species, and total number of inflorescences (F(174, 181) = 1131, p \< 0.001, alpha = 0.025). The best model was chosen by doing a model comparison with three other models using model selection and picked the model with a weight of 1 (signifies that it is the best model). As the p-value falls below the significance level, we were able to reject the null hypothesis in favor of the alternative one, which states that at least one of the factors in the data set has some influence on seed count. In a biological sense, this means that multiple factors work together to influence seed count, possibly more than the ones incorporated into this model. In this case however, plot type (shrub or open), plant species, and total number of inflorescences interact in some way to impact seed count.

## References

Bartoń K (2023). \_MuMIn: Multi-Model Inference\_. R package version 1.47.5,
 <https://CRAN.R-project.org/package=MuMIn>

Gross, K. L., & Kromer, M. L. (1986). Seed Weight Effects on Growth and Reproduction in Oenothera biennis L. *Bulletin of the Torrey Botanical Club*, *113*(3), 252--258. <https://doi.org/10.2307/2996364>

Hartig F (2022). \_DHARMa: Residual Diagnostics for Hierarchical (Multi-Level /
Mixed) Regression Models\_. R package version 0.4.6,
<https://CRAN.R-project.org/package=DHARMa>

Seaver, M. 2022. Individual and community flowering phenology, seed counts and pollinator visitation rates in shrub and open plots across Niwot Ridge, 2019 - 2021. ver 1. Environmental Data Initiative. <https://doi.org/10.6073/pasta/edc4ab2faf7dd96cd1deac1544d5f2b9>

Wickham H, Vaughan D, Girlich M (2023). \_tidyr: Tidy Messy Data\_. R package
version 1.3.0, <https://CRAN.R-project.org/package=tidyr>

Zimmerman, M. (1980). Reproduction in polemonium: Pre-dispersal seed predation. *Ecology*, *61*(3), 502--506. <https://doi.org/10.2307/1937415> 
